On multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2021-07-04 paper/mcd2.tex
Put in the basic algorithms.

\newcommand{\supp}{\operatorname{supp}}

@@

Suppose $\cV$ is a real, reflexive Banach space with norm $\|v\|$ for $v\in\cV$ and topological dual space $\cV'$.  Denote the dual pairing of $\phi \in \cV'$ and $v$ by $\ip{\phi}{v} = \phi(v)$.  Note that $\|\phi\|_{\cV'} = \sup_{\|v\|=1} |\ip{\phi}{v}|$ defines the norm on $\cV'$, which is also a Banach space.

Let $\cK \subset \cV$ be a nonempty closed and convex subset of $\cV$, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous \emph{residual functional} $f:\cK \to \cV'$ and \emph{source} $g\in \cV'$ we consider the following \emph{variational inequality} (VI) for $u\in \cK$:
\begin{equation}
\ip{f(u)}{v-u} \ge g(v-u) \qquad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is a general, nonlinear map the use of a right-hand source function is not strictly needed---by redefining $f$ we may take $g=0$---but its inclusion will be helpful in describing algorithms in Section \ref{sec:multilevel}.

Informally, if we conceptualize the dual pairing as an inner product, and if $g=0$, then \eqref{eq:vi} says that the angle between $f(u)$ and any arbitrary vector $v-u$ pointing from $u$ into $\cK$ is at most $90^\circ$, that is, $f(u)$ points directly into $\cK$ if $u$ solves \eqref{eq:vi}.  Furthermore, if $u$ is in the interior of $\cK$ then \eqref{eq:vi} implies $f(u)=g$; this VI generalizes the concept of a nonlinear system of equations $f(u)=g$ over $\cV$ to the constrained case where $\cK \subsetneq \cV$.

@@

It is well-known that if $f:\cK \to \cV'$ is continuous, monotone, and coercive then VI \eqref{eq:vi} has a solution \cite[Corollary III.1.8]{KinderlehrerStampacchia1980}, and also that the solution is unique if $f$ is strictly monotone.  The condition of continuity can be weakened to only apply on finite-dimensional subspaces \cite{KinderlehrerStampacchia1980}, but the stronger condition will apply in our examples.  As in the calculus of variations \cite{Evans2010}, coercivity permits a weak compactness argument for unbounded sets $\cK$.

@@

Our multilevel algorithms (next section) are based on both subspace and constraint decompositions, as follows.  Suppose there are $m<\infty$ subspaces $\cV_i \subset \cV$ so that
\begin{equation}
\cV = \sum_{i=1}^m \cV_i \label{eq:subspacedecomp}
\end{equation}
holds in the sense that if $w \in \cV$ then there exist $w_i \in \cV_i$ so that $w = \sum_i w_i$ \cite{Xu1992}.  Suppose further that $\cK_i \subset \cV$ are nonempty, closed, and convex subsets such that
\begin{equation}
\cK_i \subset \cV_i \qquad \text{and} \qquad \cK = \sum_{i=1}^m \cK_i. \label{eq:constraintdecomp}
\end{equation}
The constraint decomposition \cite{Tai2003} sum in \eqref{eq:constraintdecomp} is required to hold in two senses: \emph{(i)}~if $w \in \cK$ then there exist $w_i \in \cK_i$ so that $w = \sum_i w_i$, and \emph{(ii)}~if $z_i \in \cK_i$ for each $i$ then $\sum_i z_i \in \cK$.  (The latter sense is automatic in \eqref{eq:subspacedecomp} because $\cV_i$ are subspaces.)  Note that neither decomposition is unique (in general), and that $\cK_i \not\subset \cK$ in many applications.

Following \cite{Tai2003}, for each $\cK_i$ we also assume a (generally) nonlinear restriction operator $R_i : \cK \to \cK_i$ such that
\begin{equation}
v = \sum_{i=1}^m R_i v \qquad \text{if } v \in \cK.  \label{eq:constraintrestrictionsum}
\end{equation}
An \emph{constraint decomposition} of the constraint set $\cK$ is a choice of $\cV_i,\cK_i,R_i$ satisfying the above hypotheses.

@@

\begin{example}  \label{ex:domaindecomposition}  Consider a bounded domain $\Omega \subset \RR^d$ with polygonal boundary, let $\cV = H_0^k(\Omega)$, $k\ge 1$, and suppose the obstacle $\psi \in H^k(\Omega)$ satisfies $\psi|_{\partial \Omega} \le 0$.  Let $\cK = \{v \ge \psi\} \subset \cV$.  Suppose further that $\{\phi_i\}_{i=1}^m$ is a smooth partition of unity on $\Omega$, i.e.~so that $0 \le \phi_i\le 1$ and $\sum_i \phi_i = 1$, and let $\Omega_i = \supp \phi_i$ (open support).  Let $\cV_i = \{w \in \cV:w|_{\Omega \setminus \Omega_i} =0 \}$, $\cK_i = \{v \in V: v \ge \phi_i \psi\}$, and $R_i(v) = \max\{v,\phi_i \psi\}$.  Then \eqref{eq:subspacedecomp}, \eqref{eq:constraintdecomp}, and \eqref{eq:constraintrestrictionsum} all hold.
\end{example}

@@

FIXME state the basic ``sequential constraint decomposition'' and ``parallel constraint decomposition'' algorithms

In practice the space $\cV$ will be a finite-dimensional finite element (FE) space because we will solve VI \eqref{eq:vi} over a mesh.  We FIXME triangulation etc. Though our

\begin{example}  FIXME $\cV_i$ are 1d spaces on a single level; $\psi,\cK,\cK_i$ for obstacle problem; note $\cK_i \not\subset \cK$ when $\psi>0$
\end{example}

>>>>

\newcommand{\supp}{\operatorname{supp}}
\newcommand{\Span}{\operatorname{span}}

@@

Suppose $\cV$ is a real, reflexive Banach space with norm $\|v\|$ for $v\in\cV$ and topological dual space $\cV'$.  Denote the dual pairing of $\phi \in \cV'$ and $v\in\cV$ by $\ip{\phi}{v} = \phi(v)$.  Note that $\|\phi\|_{\cV'} = \sup_{\|v\|=1} |\ip{\phi}{v}|$ defines the norm on $\cV'$, which is also a Banach space.

Let $\cK \subset \cV$ be a nonempty closed and convex subset of $\cV$, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous \emph{residual functional} $f:\cK \to \cV'$ and \emph{source} $g\in \cV'$ we consider the following \emph{variational inequality} (VI) for $u\in \cK$:
\begin{equation}
\ip{f(u)}{v-u} \ge \ip{g}{v-u} \qquad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is a general, and generally nonlinear, map the inclusion of $g$ is not strictly needed---by redefining $f$ we may take $g=0$---but its inclusion will be helpful in describing algorithms in Section \ref{sec:multilevel}.

VI \eqref{eq:vi} generalizes the nonlinear system of equations $f(u)=g$ over a vector space $\cV$ to the constrained case where $\cK \subsetneq \cV$ is a convex subset.  Informally, if we conceptualize the dual pairing as an inner product, then \eqref{eq:vi} says that the angle between $f(u)-g$ and any arbitrary vector $v-u$ pointing from $u$ into $\cK$ is at most $90^\circ$.  That is, \eqref{eq:vi} says that $f(u)-g$ points directly into $\cK$ if $u\in\partial\cK$, while if $u$ is in the interior of $\cK$ then \eqref{eq:vi} implies $f(u)=g$.

@@

It is well-known that if $f:\cK \to \cV'$ is continuous, monotone, and coercive then VI \eqref{eq:vi} has a solution \cite[Corollary III.1.8]{KinderlehrerStampacchia1980}, and also that the solution is unique if $f$ is strictly monotone.  The condition of continuity can be weakened to only apply on finite-dimensional subspaces \cite{KinderlehrerStampacchia1980}, but the stronger condition will apply in our examples.  As in the calculus of variations \cite{Evans2010}, coercivity permits a compactness argument even for unbounded sets $\cK$; recall that the bounded, closed subsets of a reflexive Banach space are weakly compact.

@@

Our multilevel algorithms (next section) are based on both subspace and constraint decompositions, as follows.  Suppose there are $m<\infty$ subspaces $\cV_i \subset \cV$ so that
\begin{equation}
\cV = \sum_{i=0}^{m-1} \cV_i \label{eq:subspacedecomp}
\end{equation}
holds in the sense that if $w \in \cV$ then there exist $w_i \in \cV_i$ so that $w = \sum_i w_i$ \cite{Xu1992}.  Suppose further that $\cK_i \subset \cV$ are nonempty, closed, and convex subsets such that
\begin{equation}
\cK_i \subset \cV_i \qquad \text{and} \qquad \cK = \sum_{i=0}^{m-1} \cK_i. \label{eq:constraintdecomp}
\end{equation}
The constraint decomposition \cite{Tai2003} sum in \eqref{eq:constraintdecomp} is required to hold in two senses: \emph{(i)}~if $w \in \cK$ then there exist $w_i \in \cK_i$ so that $w = \sum_i w_i$, and \emph{(ii)}~if $z_i \in \cK_i$ for each $i$ then $\sum_i z_i \in \cK$.  (The latter sense is automatic in \eqref{eq:subspacedecomp} because $\cV_i$ are subspaces.)  Note that neither decomposition is unique (in general), and that $\cK_i \not\subset \cK$ in many applications.

Following \cite{Tai2003}, for each $\cK_i$ we also assume a (generally) nonlinear restriction operator $R_i : \cK \to \cK_i$ such that % SHOULD WE HAVE  R_i:\cV \to \cK_i?
\begin{equation}
v = \sum_{i=0}^{m-1} R_i v \qquad \text{if } v \in \cK.  \label{eq:constraintrestrictionsum}
\end{equation}
An \emph{constraint decomposition} of the constraint set $\cK$ is a choice of $\cV_i,\cK_i,R_i$ satisfying the above hypotheses.

@@

\begin{example}  \label{ex:domaindecomposition}  Consider a bounded domain $\Omega \subset \RR^d$ with polygonal boundary, let $\cV = H_0^k(\Omega)$, $k\ge 1$, and suppose the obstacle $\psi \in H^k(\Omega)$ satisfies $\psi|_{\partial \Omega} \le 0$.  Let $\cK = \{v \ge \psi\} \subset \cV$.  Suppose further that $\{\phi_i\}_{i=0}^{m-1}$ is a smooth partition of unity on $\Omega$, i.e.~so that $0 \le \phi_i\le 1$ and $\sum_i \phi_i = 1$, and let $\Omega_i = \supp \phi_i$ (open support).  Let $\cV_i = \{w \in \cV:w|_{\Omega \setminus \Omega_i} =0 \}$, $\cK_i = \{v \in V: v \ge \phi_i \psi\}$, and $R_i(v) = \max\{v,\phi_i \psi\}$.  Then \eqref{eq:subspacedecomp}, \eqref{eq:constraintdecomp}, and \eqref{eq:constraintrestrictionsum} all hold.
\end{example}

@@

\begin{pcode}[ht]
\begin{pseudo*}
\pr{cd-parallel}(f:\cK \to \cV', g\in \cV', \{\cK_i\}, \{R_i:\cK \to \cK_i\}, 0<\alpha\le 1, u\in\cK)\text{:} \\+
    for $i \in \{0,\dots,m-1\}$: \qquad \qquad \ct{parallel for} \\+
        $\hat w_i\in \cK_i$: \quad $\ip{f(u - R_i u + \hat w_i)}{v_i-\hat w_i} \ge g(v_i-\hat w_i) \quad \forall v_i\in \cK_i$ \\-
    return $w=(1-\alpha) u + \alpha \sum_i \hat w_i$
\end{pseudo*}
\caption{The basic parallel CD algorithm updates an iterate $u$ by solving VIs over each constraint set $\cK_i$ in any order, giving $w\in \cK$.}
\label{pc:cdparallel}
\end{pcode}

\begin{pcode}[ht]
\begin{pseudo*}
\pr{cd-sequential}(f:\cK \to \cV', g\in \cV', \{\cK_i\}, \{R_i:\cK \to \cK_i\}, 0<\alpha\le 1, u\in\cK)\text{:} \\+
    for $i = 0,\dots,m-1$: \qquad \qquad \ct{in order} \\+
        $\hat w_i\in \cK_i$: \quad $\displaystyle \ip{f\Big(\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j u\Big)}{v_i-\hat w_i} \ge g(v_i-\hat w_i) \quad \forall v_i\in \cK_i$ \\
        $w_i = (1-\alpha) R_i u + \alpha \hat w_i$ \\-
    return $w= \sum_i w_i = (1-\alpha) u + \alpha \sum_i \hat w_i$
\end{pseudo*}
\caption{Each VI incorporates the previous solutions in this basic sequential CD algorithm; compare Pseudocode \ref{pc:cdparallel}.}
\label{pc:cdsequential}
\end{pcode}

\begin{example}  FIXME if $\cV_i=\Span\{\phi_i\}$ for $\phi_i\in\cV$ are 1d spaces and if $\cK = \{v \ge \psi\} \subset \cV$ (obstacle problem), and if $R_i : \cK \to \cK_i$ for obstacle problem; note $\cK_i \not\subset \cK$ when $\psi>0$
\end{example}

In practice the space $\cV$ will be a finite element (FE) space because we will solve VI \eqref{eq:vi} over a mesh, and thus $\cV$ and each $\cV_i$ will be finite-dimensional.
